\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2009-12-29 honors_thesis/thesis/paper.tex
Some (final?) modifications to the GECCO paper, after my own reading through

\begin{table}
  \caption{All possible updates for a particle with two neighbors.}
  \label{tab:evals}
  \centering
  \begin{tabular}{c|c}
	$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	No update&No update\\
	Updated&No update\\
	No update&Neighbor 1\\
	No update&Neighbor 2\\
	Updated&The particle itself\\
	Updated&Neigbhor 1\\
	Updated&Neighbor 2\\
  \end{tabular}
\end{table}

Because $2(n+1)$ speculative evaluations must be performed for each particle,
the choice of topology is an important one.  The use of the Complete topology,
where every particle is a neighbor to every other particle, would require on
the order of $p^2$ speculative evaluations per iteration, where $p$ is the
swarm size.  Clearly it is much more desirable to have a sparse topology, where
$O(np)$ is much smaller than $O(p^2)$.  However, some functions are better
optimized with the Complete topology and the quick spread of information it
entails than with sparse topologies.  Accordingly, we use the Random topology
described in~\cite{mcnabb-cec09}, which has been shown to effectively simulate
the Complete topology.  In that topology, 2 neighbors are chosen at random from
the swarm for each particle at each iteration.  Because a particle is also
always its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report
the results for using both the Ring topology (where $n$ again is $3$) and the
Random topology on a number of common benchmark functions.

@@

This is done by first updating each particle $\noeval{\p}_t$ to
$\nonbest{\p}_t$ by associating the value determined by the slave to the
particle it came from, and updating that particle's personal best if the value
reported is better than its previous personal best value.  Then each
$\nonbest{\p}_t$ must be updated to $\p_t$ by updating its neighborhood best
position and value with information about each of its neighbors.  This is
simply the original PSO algorithm.

Then, with the set $\pset_t$, the branch that was actually taken by each
particle can be determined, and the correct speculative child can be chosen.
The branches depended entirely on whether or not the personal best of $p_t$ was
updated and which particle, if any, updated the particle's neighborhood best.
Both of those items are now known, and the child that matches the branch is
selected.

The parent $\p_t$ must pass its personal best value to the child, as the child
knows only the position that it guessed, not the function value at that
position.  It is possible that both $\p_t$ and $\s_{t+1}$ updated their
personal bests, but $\p_t$'s value was better.  For example, suppose that
$\p_{t-1}$ has a personal best value of 3.  $\noeval{\p}_t$ was created, and
$\noeval{\s}_{t+1}$ was moved assuming that $\p_t$ had updated its personal
best with its position at time $t$.  Then both $\noeval{\p}_t$ and
$\noeval{\s}_{t+1}$ were evaluated, with values 1 and 2, respectively.
$\nonbest{\s}_{t+1}$ would think that its current position was its personal
best, as the value it found, 2, was better than its previous personal best
value of 3.  It needs to receive the personal best value from its parent to
know that its personal best position $\pbest$ is actually the position of
$\p_t$, not $\s_{t+1}$.

@@

In a distributed parallel PSO algorithm, individual processors not only perform
evaluations of particles, but also their movement.  The information for each
particle is not held by a central machine that directs the algorithm; instead,
each processor has the information for the particle or particles that it is in
charge of and must perform the steps of the algorithm for the particles it
manages.  Messages such as values and positions for the neighborhood best are
sent between processors.  There may be some machine that collects information
from all of the particles and outputs the result of the algorithm, though that
machine's importance is much less than in centralized algorithms.

@@

To get that information, a round of message passing is required.  Each particle
sends its position to its neighbors at iteration $t$, so that all particles can
generate $\noeval{\sset}_{t+1}$.  After each particle evaluates its position
and the positions of its speculative children, it passes information about the
outcome iteration $t$ to its neighbors, so that each particle can receive
information about its neighborhood best to be updated to $\p_t$.  Once that
communication is finished, the particle can select the speculative child which
matched the branch that iteration $t$ actually produced.  Then another round of
information passing follows, for iteration $t+1$, so that $\nonbest{\p}_{t+1}$
can be updated to $\p_{t+1}$.  Two iterations have then been completed with
only one round of evaluations, and the next iteration can start again with the
first round of message passing.

@@

This second method only requires one round of passing information because
information about iteration $t$ and $t+1$ are passed at the same time.  Each
processor reconstructs from the messages it receives all of the information
that it needs about its neighbors.  Messages are passed directly after
evaluating each particle and its children, so all messages are of the form of
$\nonbest{\p}_t$ or $\nonbest{\s}_{t+1}$.  The first iteration needs to be
treated specially, so each particle can produce its initial set of speculative
children---neighbors need only pass their initial position.  This kind of
message passing necessitates the careful use of random seeds, so that when each
processor computes the motion equations for its neighbors it gets the same
results as its neighboring processors.

@@

We then need the set of neighbors to $\nonbest{\p}_{t+1}$,
$\nonbest{\nset}_{t+1}$, so we can update $\nonbest{\p}_{t+1}$'s neighborhood
best.  To produce each neighbor $\nonbest{\n}_{t+1}$, we need the same
information for the neighboring particle that we needed to produce the original
particle, $\nonbest{\p}_{t+1}$; we need the original neighbor particle, its
speculative children, and its neighbors.  With that information, the set
$\nonbest{\nset}_{t+1}$ can be obtained by following the same process used to
obtain $\nonbest{\p}_{t+1}$.  We graphically show the messages needed to
produce $\nonbest{\p}_{t+1}$ in \fig{messages3}.  Note that it looks identical
to \fig{messages2}, just with different sets of particles.

@@

This may seem like an inordinate amount of work, and in some instances it is.
However, some parallel frameworks necessitate this type of message passing, so
we showed how speculative evaluation can be performed in such frameworks.

@@

To use the extra speculative evaluations, instead of choosing the speculative
child that matches that branch that the original PSO would have taken, we take
the child that has the best value.  The methodology is exactly the same as
above except for the process of choosing which speculative child to accept.
The branch that PSO would have taken is ignored, but $\p_t$ must still be
created so it can give the right personal best and neighborhood best values and
positions to $\p_{t+1}$.  The only change needed in \alg{centralized} is in
step 7, where the $\noeval{\s}_{t+1}$ with the highest value is chosen from
$\noeval{\sset}_{t+1}$ instead of with the matching branch.

@@

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optima.  We show results in \fig{griewank1} for swarms of size 30
and 240 using the Ring topology. One can see in the figure that both SEPSO
algorithms initially outperformed PSO, but got stuck in a local optima and were
eventually outperformed.

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Instead of showing average function value, we show the
  percentage of runs that found the global optimum at each iteration.}
  \label{fig:griewank2}
\end{figure}

@@

This work represents the first attempt of which we are aware at using
additional processors in parallel with PSO to do something other than either
add particles to the swarm or parallelize the evaluation of the objective
function.  We anticipate that many future advancements can be made.  For
example, our technique required speculatively evaluating every possible next
position of each particle.  The number of speculative evaluations could
potentially be greatly reduced by using statistics for which branch of motion
particles most often follow.  Only the subset of the most probable next
locations would then be evaluated.  Using an asynchronous PSO variant would
also be helpful in this case, allowing particles to be at different iterations
in case the correct branch was not guessed and needs to be evaluated
separately.

>>>>

\begin{table}
  \caption{All possible updates for a particle with two neighbors}
  \label{tab:evals}
  \centering
  \begin{tabular}{c|c}
	$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	No update&No update\\
	Updated&No update\\
	No update&Neighbor 1\\
	No update&Neighbor 2\\
	Updated&The particle itself\\
	Updated&Neigbhor 1\\
	Updated&Neighbor 2\\
  \end{tabular}
\end{table}

Because $2(n+1)$ speculative evaluations must be performed for each particle,
the choice of topology is an important one.  The use of the Complete topology,
where every particle is a neighbor to every other particle, would require on
the order of $p^2$ speculative evaluations per iteration, where $p$ is the
swarm size.  Clearly it is much more desirable to have a sparse topology, where
$O(np)$ is much smaller than $O(p^2)$.  However, some functions are better
optimized with the Complete topology and the quick spread of information it
entails than with sparse topologies.  Accordingly, we use the Random topology
described in~\cite{mcnabb-cec09}, which has been shown to effectively simulate
the Complete topology.  In that topology, 2 neighbors are chosen at random from
the swarm for each particle at each iteration.  Because a particle is also
always its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report
the results for performing speculative evaluation using both the Ring topology
(where $n$ again is $3$) and the Random topology on a number of common
benchmark functions.

@@

This is done by first updating each particle $\noeval{\p}_t$ to
$\nonbest{\p}_t$ and each speculative particle $\noeval{\s}_{t+1}$ to
$\nonbest{\s}_{t+1}$ by associating the value determined by the slave to the
particle it came from, and updating that particle's personal best if the value
reported is better than its previous personal best value.  Then each
$\nonbest{\p}_t$ must be updated to $\p_t$ by updating its neighborhood best
position and value with information about each of its neighbors.  This is
simply the original PSO algorithm.

Then, with the set $\pset_t$, the branch that was actually taken by each
particle can be determined, and the correct speculative child can be chosen.
The branches depend entirely on whether or not the personal best of $p_t$ was
updated and which particle, if any, updated the particle's neighborhood best.
Both of those items are now known, and the child that matches the branch is
selected.

The parent $\p_t$ must pass its personal best value to the child, as the child
knows only the position that it guessed, not the function value at that
position.  It is possible that both $\p_t$ and $\s_{t+1}$ update their
personal bests, but $\p_t$'s value is better.  For example, suppose that
$\p_{t-1}$ has a personal best value of 3.  $\noeval{\p}_t$ is created, and
$\noeval{\s}_{t+1}$ is moved assuming that $\p_t$ has updated its personal
best with its position at time $t$.  Then both $\noeval{\p}_t$ and
$\noeval{\s}_{t+1}$ are evaluated, with values 1 and 2, respectively.
$\nonbest{\s}_{t+1}$ would think that its current position is its personal
best, as the value it found, 2, is better than its previous personal best
value of 3.  It needs to receive the personal best value from its parent to
know that its personal best position $\pbest$ is actually the position of
$\p_t$, not $\s_{t+1}$.

@@

In a distributed parallel PSO algorithm, individual processors not only perform
evaluations of particles, but also their movement.  The information for each
particle is not held by a central machine that directs the algorithm; instead,
each processor has the information for the particle or particles that it is in
charge of and must perform the steps of the algorithm for those particles.
Messages such as values and positions for the neighborhood best are sent
between processors.  There may be some machine that collects information from
all of the particles and outputs the result of the algorithm, though that
machine's importance is much less than in centralized algorithms.

@@

To get that information, a round of message passing is required.  Each particle
sends its position to its neighbors at iteration $t$, so that all particles can
generate $\noeval{\sset}_{t+1}$.  After each particle evaluates its position
and the positions of its speculative children, it passes information about the
outcome of iteration $t$ to its neighbors, so that neighboring particles can
update their neighborhood bests to move from $\nonbest{\p}_t$ to $\p_t$.  Once
that communication is finished, the particle can select the speculative child
which matched the branch that iteration $t$ actually produced.  Then another
round of information passing follows, for iteration $t+1$, so that
$\nonbest{\p}_{t+1}$ can be updated to $\p_{t+1}$.  Two iterations have then
been completed with only one round of evaluations, and the next iteration can
start again with the first round of message passing.

@@

This second method only requires one round of passing information because
information about iterations $t$ and $t+1$ is passed at the same time.  Each
processor reconstructs from the messages it receives all of the information
that it needs about its neighbors.  Messages are passed directly after
evaluating each particle and its children, so all messages are of the form of
$\nonbest{\p}_t$ or $\nonbest{\s}_{t+1}$.  The first iteration needs to be
treated specially, so each particle can produce its initial set of speculative
children---neighbors need only pass their initial position.  This kind of
message passing necessitates the careful use of random seeds, so that when each
processor computes the motion equations for its neighbors it gets the same
results as its neighboring processors.

@@

We then need the set of neighbors to $\nonbest{\p}_{t+1}$,
$\nonbest{\nset}_{t+1}$, so we can update $\nonbest{\p}_{t+1}$'s neighborhood
best.  To produce each neighbor $\nonbest{\n}_{t+1}$, we need the same
information for the neighboring particle that we needed to produce the original
particle, $\nonbest{\p}_{t+1}$; we need the original neighbor particle, its
speculative children, and its neighbors.  With that information, the set
$\nonbest{\nset}_{t+1}$ can be obtained by following the same process used to
obtain $\nonbest{\p}_{t+1}$.  We graphically show the messages needed to
produce $\nonbest{\nset}_{t+1}$ in \fig{messages3}.  Note that it looks
identical to \fig{messages2}, just with different sets of particles.

@@

This may seem like an inordinate amount of work, and in some instances it is.
However, some parallel frameworks necessitate this type of message passing, so
we have described how speculative evaluation can be performed in such
frameworks.

@@

To use the extra speculative evaluations, instead of choosing the speculative
child that matches that branch that the original PSO would have taken, we take
the child that has the best value.  The methodology is exactly the same as
above except for the process of choosing which speculative child to accept.
The branch that PSO would have taken is ignored, but $\p_t$ must still be
created so it can give the right personal best and neighborhood best values and
positions to $\p_{t+1}$.  The only change needed in \alg{centralized} is in
step 7, where the $\noeval{\s}_{t+1}$ with the best value is chosen from
$\noeval{\sset}_{t+1}$ instead of with the matching branch.

@@

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optima.  We show results in \fig{griewank1} for swarms of size 30
and 240 using the Ring topology. One can see in the figure that both SEPSO
algorithms initially outperformed PSO, but got stuck in a local optima and were
eventually outperformed themselves.

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Instead of showing average function value, we show the
  percentage of runs that have found the global optimum by each iteration.}
  \label{fig:griewank2}
\end{figure}

@@

This work represents the first attempt of which we are aware at using
additional processors in parallel with PSO to do something other than either
add particles to the swarm or parallelize the evaluation of the objective
function.  We anticipate that many future advancements can be made.  For
example, our technique required speculatively evaluating every possible next
position of each particle.  The number of speculative evaluations could
potentially be greatly reduced by using statistics for which branch of motion
particles most often follow.  Only the subset of the most probable next
locations would then be evaluated.  Using an asynchronous PSO variant may also
be helpful in this case, allowing particles to be at different iterations in
case the correct branch was not guessed and needs to be evaluated separately.
